---
title: "Project NFP"
author: "Sam Choi, Eric Xu"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(astsa)
```

## ARIMA

We will use the Box-Jenkins method to build an ARIMA model for the seasonally adjusted NFP data transformed using first order differencing.

### Model identification

```{r plot_ts_sa}
PAYEMS <- read.csv(file = "PAYEMS.csv", header = TRUE, sep = ",")
nfp_sa_ts_2010_2018 <- ts(PAYEMS[853:951, ][2])
SA_diff <- diff(nfp_sa_ts_2010_2018, lag = 1, differences = 1)

SA_mean <- mean(SA_diff, na.rm = TRUE)
centered_SA_diff <- SA_diff - SA_mean
plot.ts(centered_SA_diff, main = "Centered SA Diffs, 2010-2018", xlab = "Months (since January 2010)", ylab = "Number of Payrolls")
```

The time series appears stationary with a mean of approximately 188. No apparent trends.

```{r plot_acf}
acf(centered_SA_diff, main = "ACF Plot of SA Differences 2010-2018")
```

The ACF is always insignificant.

```{r plot_pacf}
pacf(centered_SA_diff, main = "PACF Plot of SA Differences 2010-2018")
```

```{r plot_ts_nsa}
PAYNSA <- read.csv(file = "PAYNSA.csv", header = TRUE, sep = ",")
nfp_nsa_ts <- ts(PAYNSA[853:951,][2])
NSA_diff <- diff(nfp_nsa_ts, lag = 1, differences = 1)

NSA_mean <- mean(NSA_diff, na.rm = TRUE)
centered_NSA_diff <- NSA_diff - NSA_mean
plot.ts(centered_NSA_diff, main = "Centered NSA Diffs, 2010-2018", xlab = "Months (since January 2010)", ylab = "Number of Payrolls")

NSA_sdiff <- diff(nfp_nsa_ts, lag = 1, differences = 1)
NSA_smean <- mean(NSA_sdiff)
centered_NSA_sdiff <- NSA_sdiff - NSA_smean
plot.ts(centered_NSA_sdiff)

NSA_sdiff2 <- diff(NSA_sdiff, lag = 12, differences = 1)
plot.ts(NSA_sdiff2)
```



```{r plot_acf_nsa}
acf(centered_NSA_diff, main = "ACF Plot of SA Differences 2010-2018")
acf(centered_NSA_sdiff, main = "ACF Plot of SA Differences 2010-2018")
acf(NSA_sdiff2)
```

The ACF is always insignificant.

```{r plot_pacf_nsa}
pacf(centered_NSA_diff, main = "PACF Plot of SA Differences 2010-2018")
pacf(centered_NSA_sdiff, main = "PACF Plot of SA Differences 2010-2018")
pacf(NSA_sdiff2)
```

Appears to be ARMA(1, 0)

### Parameter Estimation and Model Diagnostics

```{r}
sarima(centered_NSA_diff, p = 1, d = 0, q = 0, D = 1, S = 12)
sarima(NSA_sdiff2, p = 0, d = 0, q = 1)
```

### Model Selection



