---
title: "Project NFP - Spectral Analysis"
author: "Sam Choi, Eric Xu"
date: "4/11/2018"
output:
  html_document:
    fig_height: 3
    fig_width: 5
    keep_md: true
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DataComputing)
library(astsa)
#setwd("~/Dropbox/Berkeley/stat153/NFP-TSA/")
```

## Spectral Analysis

We aim to use spectral analysis to find the key frequencies of variation in the non-seasonally adjusted NFP time series.

### Raw Periodogram

We begin our spectral analysis by transforming the spectral density using a first difference and estimating the spectral density of the transformed time series using a raw periodogram. We compare this to generating a raw periodogram using the detrending process in spec.pgram.

```{r periodogram}
PAYNSA <- read.csv(file = "PAYNSA.csv", header = TRUE, sep = ",")
nfp_nsa_ts_2010_2018 <- ts(PAYNSA[853:951, ][2])
plot.ts(nfp_nsa_ts_2010_2018, main = "Non Seasonally Adjusted NFP Data")

differenced_NSA <- diff(nfp_nsa_ts_2010_2018)
plot.ts(differenced_NSA, main = "First DIfference of NSA Data")

mvspec(differenced_NSA, detrend = FALSE)
mvspec(nfp_nsa_ts_2010_2018)
```

The raw periodogram of the differenced NSA data shows 5 major peaks corresponding to cycles of approximately 12 months, 6 months, 4 months, 3 months, and 2.5 months.

Compared to the raw periodogram of the data detrended using the mvspec default, the raw periodogram of the differenced NSA data has a lower baseline around most peaks. We will compare the smoothed periodograms of the differenced and detrended data as well.

### Smoothing the Periodogram

To improve our spectral estimator, we will smooth the periodogram using various parameters to find a better estimate of the spectral density. First we use a daniell kernel with m = 1.

```{r smooth}
mvspec(differenced_NSA, kernel("daniell", 1), log = "no")
mvspec(nfp_nsa_ts_2010_2018, kernel("daniell", 1), log = "no")
```

This looks quite smooth already. Using m = 2 results in the following smoothed periodograms.

```{r smooth2}
mvspec(differenced_NSA, kernel("daniell", 2), log = "no")
mvspec(nfp_nsa_ts_2010_2018, kernel("daniell", 2), log = "no")
```

The peaks are in the same frequencies, but the smoothing has widened the peaks, possibly introducing bias. Setting m = 1 is an adequate amount of smoothing because we can identify significant peaks while minimizing bias.

Now we try the modified danielle kernel with m = 1.

```{r modified}
mvspec(differenced_NSA, kernel("modified.daniell", 1), log = "no")
mvspec(nfp_nsa_ts_2010_2018, kernel("modified.daniell", 1), log = "no")
```

This produces much sharper peaks. We can identify two major peaks: one at a frequency of about 0.17, corresponding to a semiannual cycle, and one at a frequency of about 0.33, corresponding to a quarterly cycle.

The smoothed periodogram of the detrended data has a sharp peak at 0.04 and much lower peaks in the 2 month and quarterly cycle frequency range. Based on visual inspection of the time series plot, we believe that the peaks in the quarterly cycle frequency range in particular represent key frequencies of variation, suggesting that the differenced data set produces a smoothed periodogram with more significant peaks.

There does not appear to be much bumpiness remaining so tapering should not be necessary. To check, we set taper to 0.1.

```{r taper}
mvspec(nfp_nsa_ts_2010_2018, kernel("modified.daniell", 1), log = "no", taper = 0.2)
mvspec(differenced_NSA, kernel("modified.daniell", 1), log = "no", taper = 0.2)
```

The tapering does not change the estimate spectral density significantly and the same frequency peaks remain. This suggests our data set has very clean and predictable cyclical variation.

```{r}
pgram_detrend <- mvspec(nfp_nsa_ts_2010_2018, kernel("modified.daniell", 1), log = "no")
pgram_diffed <- mvspec(differenced_NSA, kernel("modified.daniell", 1), log = "no")


key_freq_ind <- c(1, which(diff(sign(diff(pgram_detrend$spec)))==-2) + 1)
key_freq <- pgram_detrend$freq[key_freq_ind]
abline(v=key_freq, lty=2)

key_freq_ind <- c(1, which(diff(sign(diff(pgram_diffed$spec)))==-2) + 1)
key_freq <- pgram_diffed$freq[key_freq_ind]
abline(v=key_freq, lty=2)
```

Comparing with the parametric spectral estimator.

```{r}
pgram_ar <- spec.ar(differenced_NSA, plot=F)
lines(pgram_ar$freq, pgram_ar$spec, lty=2, col="red")
```

Based on our findings from the ARIMA modeling process, we believe a lagged difference with lag = 12 months and differencing again produces a stationary transformed time series that closely resembles white noise. Therefore, we concluded that fitting an ARMA model with non-zero order would not produce useful information. This suggests that parametric spectral estimation is a poor method for estimating the spectral density, because for all non-zero p, an AR(p) model would produce a poor fit and introduce a dependency structure to uncorrelated white nosie, creating bias in the estimated spectral density.

Now we use the top key frequencies to model the 

```{r}
top_freq <- c(0.17, 0.33, 0.08)
t <- 1:length(nfp_nsa_ts_2010_2018)
periodic_terms <- do.call(cbind, lapply(top_freq, function(freq) {
  cbind(cos(2 * pi * freq * t), sin(2 * pi * freq * t))
}))
df <- data.frame(nfp_nsa_ts_2010_2018, t, periodic_terms)
fit_final <- lm(nfp_nsa_ts_2010_2018 ~ ., df)
plot(t, nfp_nsa_ts_2010_2018, type="l")
lines(t, fit_final$fitted.values, lty=2, col="red")
```

### Spectral Forecasting

```{r spec_forecasting}
t_new <- (tail(t, 1) + 1):(tail(t, 1) + 24)
periodic_terms_new <- do.call(cbind, lapply(top_freq, function(freq) {
  cbind(cos(2 * pi * freq * t_new), sin(2 * pi * freq * t_new))
}))
df_new <- data.frame(t_new, periodic_terms_new)
colnames(df_new) <- colnames(df)[-1]
predictions <- predict.lm(fit_final, newdata=df_new, interval="prediction", level=.95)
plot(t, nfp_nsa_ts_2010_2018, type="l", xlim=c(0, tail(t_new, 1)), ylim=c(125000, 155000))
lines(t, fit_final$fitted.values, lty=2, col="red")
lines(t_new, predictions[, "fit"], col="blue")
matlines(t_new, predictions[, 2:3], col = "purple", lty=3)
```


