---
title: "Project NFP - Spectral Analysis"
author: "Sam Choi, Eric Xu"
date: "4/11/2018"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DataComputing)
library(astsa)
setwd("~/Dropbox/Berkeley/stat153/NFP-TSA/")
```

## Spectral Analysis

We aim to use spectral analysis to find the key frequencies of variation in the non-seasonally adjusted NFP time series.

### Raw Periodogram

We begin our spectral analysis by detrending the spectral density using a first difference and estimating the spectral density of the detrended time series using a raw periodogram.

```{r periodogram}
PAYNSA <- read.csv(file = "PAYNSA.csv", header = TRUE, sep = ",")
nfp_nsa_ts_2010_2018 <- ts(PAYNSA[853:951, ][2])
plot.ts(nfp_nsa_ts_2010_2018, main = "Non Seasonally Adjusted NFP Data")

differenced_NSA <- diff(nfp_nsa_ts_2010_2018)
plot.ts(differenced_NSA, main = "First DIfference of NSA Data")

# mvspec(nfp_nsa_ts_2010_2018)
mvspec(differenced_NSA, detrend = FALSE)
```

The raw periodogram shows 5 major peaks corresponding to cycles of approximately 12 months, 6 months, 4 months, 3 months, and 2.5 months.

### Smoothing the Periodogram

To improve our spectral estimator, we will smooth the periodogram using various parameters to find a better estimate of the spectral density. First we use a daniell kernel with m = 4.

```{r smooth}
k = kernel("daniell", 4)
# mvspec(nfp_nsa_ts_2010_2018, k, log="no")
mvspec(differenced_NSA, k, log="no")
```

The smoothed periodgram shows the greatest variance in the 0.13 to 0.21 frequency range, which corresponds to one cycle every 5-7 months, or approximately a semiannual cycle. There is also a smaller peak at 0.46, corresponding to a cycle every 2 months, and a third peak at 0.29 to 0.37, corresponding to a cycle every 2.7-3.4 months, or a quarterly cycle.

Now we set taper to 0.1. ADD REASONING.

```{r taper}
# mvspec(nfp_nsa_ts_2010_2018, k, log="no", taper = 0.1)
mvspec(differenced_NSA, k, log="no", taper = 0.1)
```

The tapering does not change the estimate spectral density significantly and the same frequency peaks remain.
